# How does calculate_range_quantiles work?

This function is hard to read because Stan is missing a lot of built-in functions that R has (like a cumulative sum) and has some other features (e.g., needing to predefine vector size) that made this tricky to write. 

Here's the whole function before we break it apart. 

```{stan}
  real calculate_range_quantile(int np, vector patches, real[] dens_by_patch, real quantile_out){
    vector[np] csum_dens;
    vector[np] csum_to_edge; 
    real quant_position; 
    real cutoff; 
    int cutoff_id; 
    
    cutoff = sum(dens_by_patch[1:np]) * quantile_out; // where along the vector of counts does the quantile cutoff fall? (then need to translate that into a patch position, below)  
    
    for(i in 1:np){
      csum_dens[i] = csum(dens_by_patch[1:np], i); // calculate cumulative sum of density along each patch 
      if(csum_dens[i] <= cutoff){
        csum_to_edge[i] = csum_dens[i]; // keep only the patches below the edge 
      } else {
        csum_to_edge[i] = 0; 
      }
    }
    
    cutoff_id = min(which_equal(csum_to_edge, 0)); // get lowest patch with a 0 (that's the patch where the edge falls)

    quant_position = ((cutoff - max(csum_to_edge)) / dens_by_patch[cutoff_id]) + cutoff_id - 1; // calculate what proportion of the edge-containing patch is "filled in" by the actual fish up to the weighted quantile (that's the decimal) and then add in the other patches 
    
    return(quant_position); 
    
  } // close quantile function
```

Let's walk through it with a toy example that looks like this. 

```{r}
patches <- seq(1, 10, 1)
dens <- c(0, 83, 571, 4438, 102, 18, 0, 7, 0, 0)
print(dens)
```

The first thing the function does is calculate a quantity called `cutoff`. This tells you at how many fish you reach the "cutoff" where the range edge should fall. For our toy example, that is: 

```{r}
cutoff_05 <- 0.05 * sum(dens)
cutoff_95 <- 0.95 * sum(dens)
cutoff_05
cutoff_95
```

This tells you at how many fish we reach the 5th and 95th percentile for the geographical distribution in this year. 

Next, we need to figure out what patch those fish are in. In other words, if we hit the 5th percentile at X fish, how do we translate that to a patch position?

This happens in two parts: we identify which integer patch it should be in (between 1 and 10), and then calculate the decimal to add on to get an exact patch "latitude". 

To get the integer patch we first create a cumulative sum vector (this function doesn't exist in Stan, unfortunately) that we call `csum_dens`. 

```{r}
csum_dens <- cumsum(dens)
csum_dens 
```

Then we need to know which integer patch is closest to the cutoff, without exceeding it. The way I figured out to do this in Stan was to create a new object, `csum_to_edge`, that counts up patches until `cutoff` is reached. For this example, `csum_to_edge` for the 0.05 quantile will be the following vector: `[0,83,0,0,0,0,0,0,0,0]` and for 0.95 it will be: `[0,83,654,0,0,0,0,0,0,0]`. Here's why:

```{r}
csum_dens <= cutoff_05
csum_dens <= cutoff_95
```

Then we define a new quantity to just pull out the single patch in which the edge falls. This is called `cutoff_id`, and it does this in Stan by looking for the first 0 in `csum_to_edge` and pulling its index value. In this example `cutoff_id` will be 3 for 0.05 and 4 for 0.95. 

Note that if only the first patch is occupied, we'll just end up with a vector of ten zeros in `csum_to_edge`. That's OK! The code will work, and identify `cutoff_id` correctly as 1 (the first patch). 

Now that we've identified the integer patch below the edge we can do the second part of the calculation which is *how far into the adjacent patch we should say the edge is sitting*. For this worked example, we can do better than just say "the equatorward edge is in patch 3 and the poleward edge is in patch 4"; we can calculate proportionally -- assuming density is continuous within each patch -- how far into the patch that edge falls. That's the last equation in the function, which just tells you proportionally how much of the next patch you would have to fill in to reach the "cutoff" number of fish we're looking for, and adds that to `cutoff_id` to get a latitudinal position. 

```{r}
(cutoff_05 - 83) / dens[3] + 3 - 1
(cutoff_95 - 654) / dens[4] + 4 - 1
```
